---
title: "Exploratory Data Analysis on Red Wine Quality"
author: "Jiayi Liao"
date: "1/20/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(GGally)
library(ggplot2)
library(gridExtra)
library(scales)
library(memisc)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
redwine <- read.csv('/Users/jiayiliao/Desktop/wineQualityReds.csv')
```

# Wine Quality Exploratory Data Analysis

This tidy data set contains 1,599 red wines with 11 variables on the chemical properties of the wine. 
At least 3 wine experts rated the quality of each wine, providing a rating between 0 (very bad) and 10 (very excellent).
This analysis includes univariate, bivariate, and multivariate analysis around which chemical properties influence the quality of red wines. 



## Univariate Plots Section

### structure of the data set
```{r echo=FALSE, message=FALSE, warning=FALSE}
str(redwine)
```

There are 1599 observations of 13 numeric variables.
`X` appears to be the index.


### statistics of the variables
```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(redwine)
```
`quality` is an ordered, categorical, discrete variable. The values ranged only from 3 to 8, with a mean of 5.6 and median of 6.
All other variables seem to be continuous quantities.
`fixed.acidity` and `volatile.acidity`,  also `free.sulfur.dioxide` and `total.sulfur.dioxide` may possible be dependent or subsets of each other.


### main features of the data set
```{r echo=FALSE, message=FALSE, warning=FALSE}
names(redwine)
```

### data set preview
```{r echo=FALSE, message=FALSE, warning=FALSE}
head(redwine)
```


### distribution of wine quality 
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(x = quality)) + geom_histogram(color = 'black', fill = 'indianred3', binwidth = 1)
```

Quality ranges from 3 to 8, with a mean between 5 and 6.


### adding new variables
```{r echo=FALSE, message=FALSE, warning=FALSE}
redwine$quality.factor <- factor(redwine$quality)
redwine$quality.cat <- NA
redwine$quality.cat <- ifelse(redwine$quality>=7, 'good', NA)
redwine$quality.cat <- ifelse(redwine$quality<=4, 'bad', redwine$quality.cat)
redwine$quality.cat <- ifelse(redwine$quality==5, 'medium', redwine$quality.cat)
redwine$quality.cat <- ifelse(redwine$quality==6, 'medium', redwine$quality.cat)

redwine$quality.cat <- factor(redwine$quality.cat, levels = c("bad", "medium", "good"))
```

Translate quality into categorial variable to look at relationship between other variables by quality grades for the analysis. Change type of the quality variable to factor and add it to the dataframe as a new variable quality.factor. In addition, 3 categories of quality was created- good (>= 7), bad (<=4), and medium (5 and 6).


### fixed acidity  
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine,aes(x = fixed.acidity))+ geom_histogram(colour='black',  fill = 'indianred3', binwidth = 0.5)
```

The distribution of fixed acidity is right skewed. The median is around 8 with high concentration of wines with Fixed Acidity but due to some outliers. 

### volatile acidity
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(x = volatile.acidity)) + geom_histogram(color = 'black', fill = 'indianred3',binwidth = 0.05) 
```

The distribution of volatile acidity has two peaks around 0.4 and 0.6.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine,
       aes(x = volatile.acidity)) +
  geom_histogram(color = 'black', fill = 'indianred3', binwidth = 0.05) +
  scale_x_log10()
```

When plotted on a base 10 logarithmic scale, `volatile.acidity` appear to be normally-distributed.

### citric acid
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = redwine, aes(x = citric.acid)) + geom_histogram(color = 'black', fill = 'indianred3', binwidth = 0.1) 
```

Apart from some outliers, the distribution of citric acid looks almost rectangular. 

### residual sugar
```{r echo=FALSE, message=FALSE, warning=FALSE}
grid.arrange(ggplot(redwine, aes( x = 1, y = residual.sugar )) + 
               geom_jitter(alpha = 0.1 ) +
               geom_boxplot(alpha = 0.2, color = 'red' ) +
               scale_y_continuous(lim = c(1,8)),
ggplot(data = redwine, aes(x = residual.sugar)) +
  geom_histogram(binwidth = 0.5, color = 'black',fill = 'indianred3') +
  scale_x_continuous(lim = c(1,8)), ncol = 2)
```

The distribution of Residual Sugar is how much sugar is left after fermentation, positively skewed with high peaks at around 2.3 with many outliers present at the higher ranges, median is around 2.3 with many outliers present at the higher ranges, 1st and 3rd quartile are within 1.9 and 2.6.

### pH
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = pH), data = redwine) + geom_histogram(colour='black', fill = 'indianred3', binwidth = 0.1)
```

pH has a normal distribution with most of the pH values between 3.2 and 3.4.

### sulphates
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = sulphates), data = redwine) + geom_histogram(colour='black', fill = 'indianred3', binwidth = 0.1)
```

Sulphates centers at 0.6, right skew all the way to 2.0
 
### alcohol
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol), data = redwine) + geom_histogram(colour='black', fill = 'indianred3', binwidth = 0.5)
```

Alcohol also follows a skewed distribution but here the skewness is less than that of Chlorides or Residual Sugars.

## Distributions of the features investigated:

- **Normal**: Quality, pH

- **Positively Skewed**: Fixed acidity, Citric acid, Sulphates, Alcohol

- **Long Tail**: Residual sugar

- **Bimodal**: Volatile acidity



# Bivariate Analysis


### relationship between wine characteristics and quality
Increasing quality of wine when chemicals levels are higher
```{r echo=FALSE, message=FALSE, warning=FALSE}
p1up = qplot(x = quality.cat, y = alcohol, fill = quality.cat,
      data = redwine,
      geom = 'boxplot')+xlab('quality')

p2up = qplot(x = quality.cat, y = sulphates, fill = quality.cat,
      data = redwine,
      geom = 'boxplot')+xlab('quality')

p3up = qplot(x = quality.cat, y = citric.acid, fill = quality.cat,
      data = redwine,
      geom = 'boxplot')+xlab('quality')

p4up = qplot(x = quality.cat, y = fixed.acidity, fill = quality.cat,
      data = redwine,
      geom = 'boxplot')+xlab('quality')

grid.arrange(p1up, p2up, p3up, p4up, ncol = 2)
```

The boxplots above show all the cases when wine quality increases with increasing values of another variables.

### decreasing value with higher wine quality 
```{r echo=FALSE, message=FALSE, warning=FALSE}
p1up = qplot(x = quality.cat, y = volatile.acidity, fill = quality.cat,
      data = redwine,
      geom = 'boxplot')+xlab('quality')

p2up = qplot(x = quality.cat, y = pH, fill = quality.cat,
      data = redwine,
      geom = 'boxplot')+xlab('quality')

p3up = qplot(x = quality.cat, y = density, fill = quality.cat,
      data = redwine,
      geom = 'boxplot')+xlab('quality')

p4up = qplot(x = quality.factor, y = density, fill = quality.cat,
      data = redwine,
      geom = "boxplot")+xlab('quality')


grid.arrange(p1up, p2up, p3up, p4up, ncol = 2)
```

Cases when wine quality decreases while the values of variables increase.

```{r echo=FALSE, message=FALSE, warning=FALSE}
pos1 <- ggplot(aes(x = fixed.acidity, y = density), data = redwine) + 
  geom_jitter(alpha = 1/5) +
  geom_smooth(method = "lm", color = "red")

pos2 <- ggplot(aes(x = fixed.acidity, y = citric.acid), data = redwine) + 
  geom_jitter(alpha = 1/5) +
  geom_smooth(method = "lm", color = "red")

grid.arrange(pos1, pos2, ncol = 2)
```

The highest positive correlation is between density and fixed.acidity, as well as between fixed.acidity.


# Multivariate Analysis and Plots


### quality classification on sulphates and alcohol 
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = log10(sulphates), y = alcohol, colour = quality.factor), 
       data = redwine) + 
  geom_point(aes(size = quality.factor)) +
  scale_color_brewer(type = 'div', palette="Dark2") +
  scale_x_continuous(lim=c(quantile(log10(redwine$sulphates), 0.01),
                           quantile(log10(redwine$sulphates), 0.99)))+
  scale_y_continuous(lim=c(quantile(redwine$alcohol, 0.01),
                           quantile(redwine$alcohol, 0.99))) 
```

Previous boxplots show increasing levels of both sulphates and alcohol increase with higher quality of red wine. The scatter plot above shows a combination of sulphates and alcohol to classify and distinguish wine quality levels.
The plot reveals a clear pattern, showing most of green and yellow dots (high-quality) in the place where both alcohol and sulphates level are high. There is also a visible range of violet dots in the middle of the plot, and orange dots(low-quality) in the bottom-left corner. This implies that such a combination of variables distinguish between different levels of wines.



# Final Plots and Summary

### plot 1
```{r echo=FALSE, message=FALSE, warning=FALSE}
p1<-
  ggplot(aes(x = fixed.acidity), data = redwine)+
        xlab("fixed acidity")+
        geom_histogram(colour="black", fill='orange', binwidth = 0.5)
p2<-
  ggplot(aes(x = volatile.acidity), data = redwine)+
        xlab("volatile acidity")+
        geom_histogram(colour="black", fill='orange', binwidth = 0.05)
p3<-
  ggplot(aes(x = citric.acid), data = redwine)+
        xlab("citric acid")+
        geom_histogram(colour="black", fill='orange', binwidth = 0.05)
p4<-
  ggplot(aes(x = residual.sugar), data = redwine)+
        xlab("residual sugar")+
        geom_histogram(colour="black", fill='orange', binwidth = 0.4)
p5<-
  ggplot(aes(x = chlorides), data = redwine)+
        geom_histogram(colour="black", fill='orange',binwidth = 0.025)
p6<-
  ggplot(aes(x = free.sulfur.dioxide), data = redwine)+
        xlab("free sulfur dioxide")+
        geom_histogram(colour="black", fill='orange', binwidth = 4)
p7<-
  ggplot(aes(x = total.sulfur.dioxide), data = redwine)+
        xlab("total sulfur dioxide")+
        geom_histogram(colour="black", fill='orange', binwidth = 20)
p8<-
  ggplot(aes(x = density), data = redwine)+
        geom_histogram(colour="black", fill='orange', binwidth = 0.001)
p9<-
  ggplot(aes(x = pH), data = redwine)+
        geom_histogram(colour="black", fill='orange', binwidth = 0.05)
p10<-
  ggplot(aes(x = sulphates), data = redwine)+
        geom_histogram(colour="black", fill='orange', binwidth = 0.1)
p11<-
  ggplot(aes(x = alcohol), data = redwine)+
        geom_histogram(colour="black", fill='orange', binwidth = 0.25)
p12<-
  ggplot(aes(x = quality), data = redwine)+
        geom_histogram(colour="black", fill='orange', binwidth = 1)
grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,p12,ncol=3)
```


### plot 2
```{r echo=FALSE, message=FALSE, warning=FALSE}
dens <- ggplot(aes(x = pH, y = density), data = redwine) + 
  geom_point(alpha = 1/5, position = position_jitter(h = 0)) +
  coord_trans(x = "log10") +
  geom_smooth(method = "lm", color = "red")

fix.ac <- ggplot(aes(x = pH, y = fixed.acidity), data = redwine) + 
  geom_point(alpha = 1/5, position = position_jitter(h = 0)) +
  coord_trans(x = "log10") +
  geom_smooth(method = "lm", color = "red")

grid.arrange(dens, fix.ac, ncol = 2)
```

### plot 3
```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = fixed.acidity, y = density, colour = quality.factor), 
       data = redwine) + 
  geom_point(size = 4) +
  #geom_point() +
  scale_color_brewer(type = 'div', palette="Set3") +
  scale_x_continuous(lim=c(quantile(redwine$fixed.acidity, 0.01),
                           quantile(redwine$fixed.acidity, 0.99))) +
  scale_y_continuous(lim=c(quantile(redwine$density, 0.01),
                           quantile(redwine$density, 0.99)))
```

## Summary

- Quality, pH have normal distribution.
- Fixed acidity, citric acid, sulphates, alcohol are positively skewed.
- The highest positive correlation is between density and fixed acidity.
- Key factors that determine and drive wine quality are alcohol, sulphates, and acidity.  

### Reflections

In this exercise, my main struggle was to get as few errors as possible. This requires lots of searching on Google, reviewing the lectures, and other research. I feel more confident using R Studio after this exericise and I'm looking forward to do more analysis on R in the future. 